\documentclass[11pt, reqno]{amsart}

\input{prelude.tex}

\usepackage[backend=biber, style=alphabetic]{biblatex}
\usepackage{microtype}
\addbibresource{bibliography.bib}

\title{Highest Weight Vectors in $\Sym(\fg)$}
\author{Patrick Norton \and Elijah Whitlam-Sandler \and Pranay Pingali\\
with Guangyi Zhang}

\begin{document}
\thispagestyle{empty}
\setcounter{page}{0}
\topskip0pt
\vspace*{\fill}
\begin{center}
  Highest Weight Vectors in $\Sym(\fg)$\\[10pt]
  Elijah Whitlam-Sandler\\
  \href{mailto:elijahw@reed.edu}{elijahw@reed.edu}\\[10pt]
  with\\[10pt]
  Zajj Daugherty\\
  \href{mailto:zdaugherty@reed.edu}{zdaugherty@reed.edu}\\[10pt]
  and\\[10pt]
  Noah Charles\\
  \href{mailto:ncharles@reed.edu}{ncharles@reed.edu}
\end{center}
\vspace*{\fill}
\newpage

\begin{abstract}
  Lie Algebras, since their discovery, have grown to become an integral link between mathematics and theoretical physics. The highest weight vectors of a Lie algebra $\fg$ are key in describing the structure of $\fg$, but are also important objects in this link between mathematics and physics, as they \emph{are} the eigenvectors of the Laplace-Casimir operator. There is not currently an easy way to determing the highest weight vectors of $\Sym(\fg)$, which limits the incorporation of Lie theory into theoretical physics. We aim to provide a classification of the heighest weight vectors of $\Sym(\fg)$ for $\fg=\fsl_n,\fso_n$, and $\fsp_n$, and to create a way to computationally determine such vectors so they can be used efficiently in physics applications.
\end{abstract}


% \begin{abstract}
%   % TODO
%   This project is an attempt to classify the structure of certain mathematical
%   objects. % TODO: Explain what mathematical objects
%   The structure of these objects relates to important questions in both
%   physics and mathematics. % TODO: What questions?
%   We know that the structure of Lie algebras is completely controlled by
%   specific elements, known as highest weight vectors. Our goal is to find all
%   these highest weight vectors and determine how they relate to each other.
%   % TODO: Prior art
% \end{abstract}

\maketitle

\section{Background and Rationale}




This project is in mathematical physics and representation theory. Specifically, we will be studying certain polynomials and their role in continuous phase transitions. A \emph{continuous phase transition} is a change to a system where some metric describing the system changes continuously over the transition \cite[\S1]{Sa06}. This is opposed to the general use of a phase transition where the phase changes may happen discretely. The classic example of a continuous phase transition is studying the magnetization---the strength of a magnet per unit volume---of a material as the material is heated. Applied heat causes the magnetization of a material to decrease continuously until the material is no longer magnetized. This is in contrast to a dis-continuous phase transition such as the boiling of water, where the density of the water changes discontinuously as the water changes from a liquid state to a gas state.

If we consider disordered electronic systems in metals (systems where the arrangement of the ions is not perfectly regular \cite{dicastro2006disordered}), the electron states can be either all localized or all delocalized depending on the properties of the system. \emph{Anderson transitions} describe the transitions between these two phases \cite{Evers_2008}. To study Anderson transitions, we can consider the ``Renormalization group action'' (RG action) on the states (formally, the states are maps $Q : M \to X$ where $M, X$ are manifolds and $X$ is called the target space; the RG action is on the space of functions from $X$ to $R$): this action consists of a procedure whereby we assign sets of points average values, and then rescale to obtain a state of the same ``size'' again. This gives us information about what properties of the system are preserved under changes of scale.

Anderson transitions have various symmetries (simple examples include symmetry under rotations or time-reversal); these symmetries can be described mathematically as Lie groups under which the Lagrangian is invariant. The target space is now naturally described by a ``Riemannian symmetric space'' $G/K$, where $K$ is a (not necessarily normal) subgroup, and the RG action is given by a differential operator that is also invariant under this group. The problem of finding physical characterizations of the state that are scale-invariant can now be restated as the mathematical problem of finding eigenfunctions for a differential operator invariant under a Lie group $G$. From a result in representation theory, we know that this differential operator is a polynomial in the ``Laplace-Beltrami'' operator, and its eigenfunctions are the highest-weight vectors of the Lie algebra $\fg$ associated with $G$. 

A Lie algebra $\fg$ (pronounced ``Lee'') is a vector space, together with a non-associative ``multiplication'' called a \emph{Lie bracket}
\[[,]: \mathfrak{g} \times \mathfrak{g} \rightarrow \mathfrak{g}.\]
To each Lie algebra, we associate a set of polynomials, denoted $\Sym(\fg)$. This set is a special example of a Lie algebra \emph{module}, meaning that it is a vector space equipped with an \emph{action} by $\fg$: the elements of $\fg$ define linear functions from $\Sym(\fg)$ to itself. In this project, instead of studying all of $\fg$, we narrow our focus to only studying highest weight vectors in $\Sym(\fg)$, which is done without compromising the physical meaning, as the highest weight vectors in $\Sym(\fg)$ can be mapped back to the original function space.

Generally, a \emph{weight vector} is an element $v$ of a module such that everything that acts on $v$ scales it by some constant factor. It turns out that being able to describe these weight vectors is sufficient to be able to completely describe the submodules of any module. A \emph{highest weight vector} is a weight vector that is in the kernel of specific actions. Weight vectors are encoded in $\fg$-modules, which can be computationally difficult to find, so we instead use some tools from Lie theory. Each Lie algebra has a large family of pairwise commuting operators in $\fg$, called the \emph{Cartan} subalgebra, whose eigenvalues reveal the structure of any (reasonably nice) $\fg$-module. With this, our primary goal is to
\begin{center}
  \fbox{compute the eigenvalues and eigenvectors of the Cartan in $\Sym(\fg)$.}
\end{center}

% Our goal is to find those specific polynomials that generate the fundamental building blocks of our favorite $\fg$-module. 

% Similarly to how numbers decompose into primes, we can decompose Lie algebras into subalgebras, and just like factoring large numbers is computationally difficult, we need special tools for decomposing large $\fg$-modules into their simple pieces.


% Like how algebras break down into subalgebras, we can break down modules into
% submodules. This is the overall goal of our project: to break down $\Sym(\mathfrak{g})$
% into submodules.

% In addition to building up with Lie algebras, we can also break them down.

%
%A subalgebra of particular importance is the Cartan
%subalgebra. Large family of pairwise commuting operators in $\fg$, whose eigenvalues reveal the structure of any (reasonably nice) $\fg$-module. 
%
%
%An important structure related to Lie algebras is that of the Lie module. A Lie
%module is a vector space that is ``acted on'' by a Lie algebra---that is, each
%element of the algebra induces some transformation on the vector space in a
%particularly nice way. This project focuses on one specific module, which we
%notate $\Sym(\mathfrak{g})$. 
%


% TODO:
% Lie algebra
% Cartan subalgebra
% Weight vectors
% Highest weight vectors
% Modules & highest weight modules
% Relevant theorems (with citations)
% Allegedly, we know what the weights are in Sym(g)---what are they (w/source)
% All that physics stuff

\section{Specific Aim and Hypothesis}

This project aims to build on the theory regarding the highest weight vectors in $\Sym(\fg)$ by providing a classification of the highest weight vectors. Assuming that $\fg$ is finite dimensional, $\Sym(\mathfrak{g})\coloneqq F\left[ \cB \right]$ is a polynomial ring where $F$ is the underlying field of $\fg$ and $\cB$ is a basis of $\fg$. As a starting point, some work has been done on what the weights of $\Sym(\fg)$ are, and there has been additional work done on decomposing $\Sym(\fg)$ into invariant $\fg$-modules $\Sym(\fg)^\fg$ and a vector space $\cH$ of harmonic polynomials \cite[(3.10)]{NeRa2004}
\[\Sym(\fg)\cong\Sym(\fg)^\fg\otimes\cH.\] 

Our secondary goal is to be able to build bridges between representation theory, mathematical physics, and computational methods. There are many interesting results in representation theory that are not necessarily known to many physicists, and this can result in duplicated effort between the fields.

% \begin{itemize}
% \item Primary: Compute hw vectors
% \item Secondary: Build bridges between deeper representation theory literature/results, mathematical physics, and computational methods


% \end{itemize}

% ***This section needs work, I'm not sure what hypotheses can be made considering that we're not doing an experiment, and I'm not sure how much further we can go into the specific aim, maybe discuss what is already known and some possible first steps, although that could also go in the next section***
% TODO:
% Explain what "classify" means

\section{Design and Procedure}

We'll begin by unpacking high level theorems about the decomposition of $\Sym(\fg)$ stated for general Lie algebras in terms of the three key examples pertinent to our study: the ``special linear'', ``special orthogonal'', and ``symplectic'' Lie algebras, notated $\fsl_n$, $\fso_n$, and $\fsp_n$ respectively. This will give us an opportunity to allow us to see the decomposition happen, and give ideas as to possible classifications and help us determine how to proceed.

Methods for decomposing semisimple finite-dimensional Lie algebras (such as $\fsl_n$) are known \cite{Dau14}, but only really provide an understanding of what is happening at a higher level. The decompositions do not provide concrete highest weight vectors that are required for Lie theory to be translated to physics. Because so much structure is used when working with Lie algebras at a high level, our goal is to find ways to exploit the known structure, and make use of it at a lower level.

% TODO: Where in Dau14?
% As a base to work from, \cite{Dau14} gives us a high-level decomposition into weight spaces. However, this method is not concrete enough to be useful from a physics perspective.

In addition to that, we plan to review the infinite families of weight vectors appearing in \cite{C20}. % TODO: Cite

Once we have some idea of the structure, we plan to then write code to derive all highest weight vectors of small degree. Having generated lots of examples, we should be able to use projection operators to build highest weight vectors in a more generalizable way. Once we have these, we should be able to leverage these results to build highest weight vectors in general degree. We plan to limit ourselves to vectors specific to the Lie algebras of types $A_{n}$, $C_{n}$, and $D_{n}$ because these are the types most applicable to continuous phase transitions.

% \begin{itemize}
% \item Do the first few small examples, seeing the decomposition happen for $\fg = \fsl_n$ ($n=3,4$), $\fsp_{6}$, $\fso_6$ in small degree homogeneous components.
% \item Review infinite families of weight vectors appearing in NOAH's THESIS.
% \item Generate code to derive all highest weight vectors in small degree.
% \item Use projection operators to build highest weight vectors in a more sophisticated way. (We know the weights)
% \item Leverage these results to build hw vecs in general degree (specific to types ACD)
% \end{itemize}

\section{Predicted and Alternative Outcomes}

Our predicted outcome is to find either recursive or closed formulas for the highest weight vectors of $\Sym(\fg)$, which will take the form of polynomials as $\Sym(\fg)$ is a polynomial ring. In the chance we are not able to achieve this goal, there are many meaningful contributions that can be made to the literature.

Much of the Lie theory literature that exists does so at a very high level, and offers little to no insight about the applications of such theory. There is great value in being able to ground the currently available high level Lie theory so that it is more easily accessible to those without a background in representation theory who want to exploit the results currently out of reach.

It would be equally useful to add context to some of the current literature that is only truly available to experts in the field. There are many `results' that could be more accurately referred to as unwritten corollaries, which adding context to would allow the material to be used at a lower level.

There is also the outcome where we create computational methods based on previous theory. This would help increase the accessibility of the work on Lie algebras by making it easier for researchers in various fields to create working examples without necessarily needing to acquire a deep knowledge of the theory behind it.


\section{Role of Student}

The role of the students in this project will be to do computations, create examples, and prove theorems in order to build up to a classification of the highest weight vectors in $\Sym(\fg)$. The students will work together to publish a paper highlighting either a discovered classification, or any progress made towards such a result. 

This project requires background knowledge of abstract algebra, and more specifically Lie theory. Patrick, Elijah, and Pranay all have experience with groups, rings, and modules from taking Math 332 (\emph{Abstract Algebra}), and Patrick and Elijah have further experience with Lie algebras from the Fall 2023 section of Math 412 (\emph{Topics in Algebra: Lie Algebras}). Patrick and Elijah are also pursuing a Math-CS interdisciplinary major and a CS minor, respectively, which will aid them in writing code that can efficiently determine the highest weight vectors of a Lie algebra $\fg$. Pranay is pursuing a physics major that will allow them to help draw meaningful connections between high level math and theoretical physics.


\section{Role of Faculty}

Zajj Daugherty and Noah Charles will help by being able to provide a high-level view of the project, as well as being able to direct a literature search. In addition, they will meet with the students multiple times per week in order to help provide structure to the project. This will also help them in providing a mechanism for accountability and giving a timetable for research. They can also function as sounding boards for inquiry and give advice as questions come up.

\section{Benefit to Student}

This project is an exciting opportunity for students to collaborate across three disciplines: physics, mathematics, and computer science. The project is driven by questions in physics, relies on mathematical machinery, and will benefit strongly from computational expertise. Though each student brings a different background and skill set, each member of the team will be building part of the interdisciplinary bridge, learning how familiar topics expand into nearby fields of study. 

From the physical perspective, we are motivated by a problem in scaling theory. The physicists on the team (Guangyi and Pranay) will derive the form of the problem that lends itself to a representation theoretic approach. In the process, they will learn techniques in Lagrangian mechanics, Gaussian integration in physics, phase transitions and the renormalization group approach, and symmetry group actions in statistical field theory. These techniques are foundational across mathematical physics and will help students transition into future projects across a broad range of subdisciplines. Both physics majors will have taken abstract algebra by this summer, and will gain further experience applying algebraic techniques.

From the mathematical perspective, the question transforms into one of studying particular representations of Lie algebras. The mathematicians on the team (Patrick and Elijah) have taken a first course in Lie theory, and this project launches quite nicely from where that course left off. Students will learn how to ground some of the larger theorems of Lie theory in specific examples and computational methods, as well as how the mathematical structure translates back into physical applications. 

Specifically, ...
\printbibliography

\end{document}
